Ride at Your Own Risk: Insights into Toronto’s Bicycle Theft Hotspots

### 2024-11-10 paper/paper.Rmd
Add relevant plots and animations

---
title: "Ride at Your Own Risk: Insights into Toronto’s Bicycle Theft Hotspots"
subtitle: "[To Be Updated....]"
author: "Yingke He, Ziheng Zhong"
thanks: "Code and data are available at: [https://github.com/iJustinn/Toronto_Bicycle_Thefts/tree/main](https://github.com/iJustinn/Toronto_Bicycle_Thefts/tree/main)."
date: today
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
bibliography: references.bib
---

@@

# you are welcome to use either the CSV file or the GeoJSON file
# GeoJSON format maybe useful if you would like to combine
# an external geospatial dataset
bike_thefts_csv <- read_csv(here("data/Bicycle_Thefts_Open_Data.csv")) |>
  mutate(
    occurence_date = as_date(
      OCC_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    report_date = as_date(
      REPORT_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    )
  )
bike_thefts_sf <- read_sf(here("data/Bicycle_Thefts_Open_Data.geojson")) |>
  mutate(
    occurence_date = as_date(
      OCC_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    report_date = as_date(
      REPORT_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    )
  )
```
Bar chart
```{r, include = TRUE, warning = FALSE, message = FALSE}
# Load the dataset
bike_thefts <- read_csv(here("data/Bicycle_Thefts_Open_Data.csv"))

# Create a bar chart for the number of bicycle thefts by month
bike_thefts %>%
  count(OCC_MONTH) %>%
  ggplot(aes(x = reorder(OCC_MONTH, -n), y = n, fill = OCC_MONTH)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Bicycle Thefts by Month", x = "Month", y = "Number of Thefts") +
  theme_minimal()

# Save the plot
# ggsave(here("other/charts/thefts_by_month.png"))
```
Pie chart
```{r, include = TRUE, warning = FALSE, message = FALSE}
# Create a pie chart for the status of bicycle thefts
bike_thefts %>%
  count(STATUS) %>%
  ggplot(aes(x = "", y = n, fill = STATUS)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  labs(title = "Bicycle Theft Status (Stolen vs Recovered)") +
  theme_minimal()
```
Bar chart
```{r, include = TRUE, warning = FALSE, message = FALSE}
# Create a bar chart for bicycle thefts by day of the week
bike_thefts %>%
  count(OCC_DOW) %>%
  ggplot(aes(x = reorder(OCC_DOW, -n), y = n, fill = OCC_DOW)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Bicycle Thefts by Day of the Week", x = "Day of the Week", y = "Number of Thefts") +
  theme_minimal()
```
Map
```{r, include = TRUE, warning = FALSE, message = FALSE}
# Load cleaned coordinates data
coordinates_data <- read.csv(here("data/cleaned_coordinates_data.csv"))

@@

# When Are Bikes at the Greatest Risk? A Look at Seasonal and weekly Patterns
Explore temporal insights, such as trends by month, day of the week with visualizations that reveal when thefts are most likely to occur.

@@

# A Decade of Disappearing Bikes: Tracing Theft Trends Over Ten Years
- timeline

>>>>

---
title: "Ride at Your Own Risk: Insights into Toronto’s Bicycle Theft Hotspots"
subtitle: "[To Be Updated....]"
author: "Yingke He, Ziheng Zhong"
thanks: "Code and data are available at: [https://github.com/iJustinn/Toronto_Bicycle_Thefts/tree/main](https://github.com/iJustinn/Toronto_Bicycle_Thefts/tree/main)."
date: today
output: html_document
bibliography: references.bib
---

<style>
body {
  background-color: #f0f8ff; /* AliceBlue color */
}
</style>

@@

# you are welcome to use either the CSV file or the GeoJSON file
# GeoJSON format maybe useful if you would like to combine
# an external geospatial dataset
bike_thefts_csv <- read_csv(here("data/clean_data.csv")) |>
  mutate(
    occurence_date = as_date(
      OCC_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    report_date = as_date(
      REPORT_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    )
  )
bike_thefts_sf <- read_sf(here("data/Bicycle_Thefts_Open_Data.geojson")) |>
  mutate(
    occurence_date = as_date(
      OCC_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    report_date = as_date(
      REPORT_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    )
  )
```
Map
```{r, include = TRUE, warning = FALSE, message = FALSE}
# Load cleaned coordinates data
coordinates_data <- read.csv(here("data/cleaned_coordinates_data.csv"))

@@

# When Are Bikes at the Greatest Risk? A Look at Seasonal and weekly Patterns

```{r mothly_plot, include = TRUE, warning = FALSE, message = FALSE, fig.width=11, fig.height=5}
# Load necessary libraries
library(tidyverse)
library(ggridges)
library(cowplot)
library(here)

# Load your data
data <- read_csv(here("data", "Theft_and_Found_Counts_By_Month_2014_2024.csv"), show_col_types = FALSE)

# Convert month names to month numbers for ordering
data <- data %>%
  mutate(OCC_MONTH_NUM = factor(OCC_MONTH, levels = month.name))

# Specify levels for OCC_YEAR to order from 2014 to 2024
data$OCC_YEAR <- factor(data$OCC_YEAR, levels = 2014:2024)

# Calculate a scaled height for "Found" data by dividing by the maximum "Found" value
max_found <- max(data$Found, na.rm = TRUE)
max_stolen <- max(data$Stolen, na.rm = TRUE)

# Plot for Bikes Found with scaled heights
plot_found <- ggplot(data, aes(x = OCC_MONTH_NUM, y = OCC_YEAR, 
                               height = (Found / max_found) * 4, group = OCC_YEAR)) +
  geom_ridgeline(
    fill = "#05595B", color = "white", alpha = 0.5, scale = 1, linewidth = 0
  ) +
  labs(y = "", x = NULL) +  # Remove x-axis title
  theme_minimal() +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(size = 0.5),
    panel.border = element_blank(),
    plot.title = element_text(size = 10, color = "#05595B", hjust = -0.1, vjust = 1.5),  # Smaller title
    panel.background = element_rect(fill = "#f0f8ff", color = NA),  # Light blue background for plot area
    plot.background = element_rect(fill = "#f0f8ff", color = NA)    # Light blue background for overall plot
  ) +
  scale_x_discrete(labels = month.abb) +
  ggtitle("Bikes Found")  # Add title for Bikes Found

# Plot for Bikes Stolen with scaled heights and no borders
plot_stolen <- ggplot(data, aes(x = OCC_MONTH_NUM, y = OCC_YEAR, 
                                height = (Stolen / max_stolen) * 4, group = OCC_YEAR)) +
  geom_ridgeline(
    fill = "#603601", color = "white", alpha = 0.5, scale = 1, linewidth = 0
  ) +
  labs(y = "", x = NULL) +  # Remove x-axis title
  theme_minimal() +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(size = 0.5),
    panel.border = element_blank(),
    panel.background = element_rect(fill = "#f0f8ff", color = NA),  # Light blue background for plot area
    plot.background = element_rect(fill = "#f0f8ff", color = NA),   # Light blue background for overall plot
    plot.title = element_text(size = 10, color = "#603601", hjust = -0.1, vjust = 1.5)  # Smaller title
  ) +
  scale_x_discrete(labels = month.abb) +
  ggtitle("Bikes Stolen")  # Add title for Bikes Stolen

# Combine the plots side by side
combined_plot <- plot_grid(plot_found, plot_stolen, nrow = 1)

# Display the combined plot
combined_plot

```

```{r interactive_weekly_plot, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=7, fig.height=4}
library(plotly)
library(readr)
library(dplyr)
library(here)

# Load the data
data <- read_csv(here("data", "Theft_Counts_By_Day.csv"), show_col_types = FALSE)

# Arrange data by day of the week (Monday to Sunday)
data <- data %>%
  arrange(factor(OCC_DOW, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")))

# Initialize plot with a line trace for the entire data
fig <- plot_ly(data, x = ~OCC_DOW, y = ~Number_of_Thefts, type = 'scatter', mode = 'lines+markers',
               line = list(color = 'brown'), marker = list(color = 'blue', size = 8))

# Add points for each day with initial visibility settings
for (i in 1:nrow(data)) {
  fig <- fig %>%
    add_trace(
      x = data$OCC_DOW[i],
      y = data$Number_of_Thefts[i],
      type = 'scatter',
      mode = 'markers+text',
      text = paste("Thefts:", data$Number_of_Thefts[i]),
      textposition = "top center",
      marker = list(color = "yellow", size = 10),
      showlegend = FALSE,
      visible = ifelse(i == 1, TRUE, FALSE)
    )
}

# Define steps for each day in the slider
steps <- lapply(1:nrow(data), function(i) {
  step <- list(
    method = "restyle",
    args = list("visible", as.list(rep(FALSE, nrow(data) + 1))),  # +1 to include the main line trace
    label = data$OCC_DOW[i]
  )
  step$args[[2]][i + 1] <- TRUE  # Make only the current day’s point visible (offset by 1 due to main line trace)
  step$args[[2]][1] <- TRUE       # Keep the main line trace visible
  return(step)
})

# Add layout with the slider and set background colors
fig <- fig %>%
  layout(
    xaxis = list(title = ""),
    yaxis = list(title = "Number of Thefts"),
    sliders = list(
      list(
        active = 0,
        currentvalue = list(prefix = "Day: "),
        steps = steps
      )
    ),
    plot_bgcolor = "#f0f8ff",     # Set background color for the plotting area
    paper_bgcolor = "#f0f8ff"     # Set background color for the entire plot
  )

# Display the plot
fig

```

@@

# A Decade of Disappearing Bikes: Tracing Theft Trends Over Ten Years

```{r animated_bar_plot, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=7, fig.height=4}
# Libraries
library(ggplot2)
library(gganimate)
library(readr)
library(here)
library(dplyr)

# Load the data
data <- read_csv(here("data", "Stolen_Found_Counts.csv"), show_col_types = FALSE)

# Rename "THEFT" to "Stolen" in the Offence_Category column
data <- data %>%
  mutate(Offence_Category = ifelse(Offence_Category == "THEFT", "Stolen", Offence_Category))

# Set the number of frames for the animation
max_frames <- 100  # Total number of frames for smooth animation

# Generate interpolated data for simultaneous increase
animated_data <- data %>%
  # Repeat each category for each frame
  group_by(Offence_Category) %>%
  do(data.frame(
    Offence_Category = .$Offence_Category,
    Number_of_Offences = seq(0, .$Number_of_Offences, length.out = max_frames),
    frame = 1:max_frames
  ))

# Ensure Offence_Category is a factor for proper ordering
animated_data$Offence_Category <- as.factor(animated_data$Offence_Category)

# Define custom colors for each category
custom_colors <- c("Stolen" = "#603601", "FOUND" = "#4682B4", "Other" = "#eedaaf")

# Create the animated bar plot
ggplot(animated_data, aes(x=Offence_Category, y=Number_of_Offences, fill=Offence_Category)) + 
  geom_bar(stat='identity') +
  theme_minimal() +
  labs(x = 'Offense Category', y = 'Number of Offences') +

  # Apply custom colors
  scale_fill_manual(values = custom_colors) +

  # Remove the legend title
  guides(fill = guide_legend(title = NULL)) +

  # Set background colors for panel and plot
  theme(
    panel.background = element_rect(fill = "#f0f8ff", color = NA),
    plot.background = element_rect(fill = "#f0f8ff", color = NA)
  ) +

  # gganimate specific settings for smooth transition
  transition_states(
    frame,
    transition_length = 2,
    state_length = 1
  ) +
  ease_aes('linear')  # Ensures a smooth, even increase

```
<div style="color: grey; font-size: 12 px; margin-top: 3px;"> Displays the count of various bicycle-related offences reported to the Toronto Police Service, highlighting categories such as stolen, found, and other incidents. The smooth increase in bar heights emphasizes the scale of each offence type, offering insights into the distribution and relative frequency of these incidents from 2014 to 2024 </div> \

```{r animated_line_plot, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=6, fig.height=3}
# Load necessary libraries
library(ggplot2)
library(gganimate)
library(dplyr)
library(hrbrthemes)
library(readr)
library(here)

# Load your data
data <- read_csv(here("data", "Bicycle_Theft_Monthly_Counts.csv"), show_col_types = FALSE)

# Create a numeric month order variable
data <- data %>%
  mutate(
    OCC_MONTH_NUM = match(OCC_MONTH, c("January", "February", "March", "April", "May", 
                                       "June", "July", "August", "September", "October", 
                                       "November", "December"))
  )

# Create the animated plot
plot <- data %>%
  ggplot(aes(x = OCC_MONTH_NUM, y = Number_of_Thefts, group = as.factor(OCC_YEAR), color = as.factor(OCC_YEAR))) +
  geom_line() +
  geom_point() +
  scale_color_viridis_d() +
  ylab("Number of Bikes Stolen") +
  xlab("Month") +
  scale_x_continuous(breaks = 1:12, labels = c("January", "February", "March", "April", "May", "June", 
                                               "July", "August", "September", "October", "November", "December")) +
  theme_ipsum() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.background = element_rect(fill = "#f0f8ff", color = NA),
    plot.background = element_rect(fill = "#f0f8ff", color = NA)
  ) +
  transition_reveal(OCC_MONTH_NUM) +
  labs(color = "Year")

# Render the animation
animate(plot, nframes = 100, fps = 10)

```
<div style="color: grey; font-size: 12 px; margin-top: 3px;"> Visualizes the monthly count of bicycle thefts reported to the Toronto Police Service from 2014 to June 2024. The animation reveals seasonal patterns and variations across different years, providing insights into peak months and potential trends over time. </div> \
